High-level cognition is supported by information-rich but compressible brain activity patterns

### 2023-12-23 paper/main.tex
finished addressing "writing-based" comments from reviewers.  still to do: control analysis with network labels...i'm trying to figure out how to do this in a way that will actually run in a reasonable amount of time...

Several recent studies suggest that the complexity of brain activity is
task-dependent, whereby simpler tasks with lower cognitive demands are
reflected by simpler and more compressible brain activity patterns, and more
complex tasks with higher cognitive demands are reflected by more complex and
less compressible brain activity patterns~\citep{MackEtal20, OwenEtal21}. These
findings complement other work suggesting that functional connectivity
(correlation) patterns are task-dependent~\citep{FinnEtal17, OwenEtal20,
ColeEtal14}, although see~\cite{GratEtal18}. Higher-order cognitive processing
of a common stimulus also appears to drive more stereotyped task-related
activity and functional connectivity across individuals~\citep{HassEtal08,
LernEtal11, SimoChan20, SimoEtal16}.

The above studies are consistent with two potential descriptions of how
cognitive processes are reflected in brain activity patterns. One possibility
is that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns increases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings, or compressed representations of brain
recordings, should carry more information during complex or high-level (versus
simple or low-level) cognitive tasks.

@@

>>>>

Several recent studies suggest that the complexity of brain activity is
task-dependent, whereby simpler tasks with lower cognitive demands are
reflected by simpler and more compressible brain activity patterns, and more
complex tasks with higher cognitive demands are reflected by more complex and
less compressible brain activity patterns~\citep{MackEtal20, OwenEtal21}. These
patterns hold even when the stimulus itself is held
constant~\citep{MackEtal20}. These findings complement other work suggesting
that functional connectivity (correlation) patterns are
task-dependent~\citep{FinnEtal17, OwenEtal20, ColeEtal14}, although
see~\cite{GratEtal18}. Higher-order cognitive processing of a common stimulus
also appears to drive more stereotyped task-related activity and functional
connectivity across individuals~\citep{HassEtal08, LernEtal11, SimoChan20,
SimoEtal16}.

The above studies are consistent with two potential descriptions of how
cognitive processes are reflected in brain activity patterns. One possibility
is that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns decreases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings should carry more information (over and
above the information carried by other features) during complex or high-level
(versus simple or low-level) cognitive tasks. The tradeoffs between these two
aspects of brain activity may also vary across brain regions or networks, for
example according to each region's functional role.

@@

One potential limitation of our work concerns how our measure of
informativeness might generalize across tasks or cognitive representations and
processes. Our use of across-participant temporal decoding accuracy as a proxy
for informativeness is motivated in part by prior work that introduced
across-participant similarity (in time-varying response to a stimulus) as a
means of identifying stimulus-driven brain activity
patterns~\citep{SimoEtal16}. Intuitively, only activity patterns are are driven
by the stimulus would be expected to synchronize (i.e., be time-locked to the
stimulus) across participants. This approach implicitly removes idiosyncratic
responses (e.g., neural patterns that are \textit{not} similar across people).
However, there are also some published examples, including in our own prior
work, that indicate that some types of stimulus-evoked activity will be missed
by across-participant comparisons. For example we have reported how brain
regions like the ventromedial prefrontal cortex (vmPFC) show stimulus-driven
responses that are, for the most part, not similar across people. In that paper
(and drawing on other work), we suggest that the vmPFC seems to represent or
support highly idiosyncratic internal states, like affective responses.
Although we would consider the vmPFC to be a ``high-level'' region (e.g., we
consider affect to be a relatively high-level aspect of cognition), the measure
of informativeness that we used in our current study would identify regions
like the vmPFC as having \textit{low} informativeness. This is because
across-participant decoding accuracy (our proxy measure for informativeness)
will only be high for representations or responses that are common across
people.

Relatedly, even in the experimental conditions we describe as ``less
cognitively engaging,'' we do not think it likely that high-level thought or
cognitive processing is absent. Rather, we suggest that these high-level
representations tend to be more idiosyncratic when the stimulus is less
engaging, and therefore less constraining on people's thoughts. Nonetheless,
even during highly engaging tasks, people may engage in idiosyncratic
stimulus-driven processes. For example, people might retrieve personal
information as they listened to the story. Those retrievals could happen at
different times for different people according to each individual's prior
experiences, and even when those sorts of retrievals happen to be temporally
synchronized across people, the specific memories or information being
retrieved might still be idiosyncratic. Our measure of informativeness is
insensitive to these processes. Further, even in response to an identical
stimulus, task instructions or participants' interal goals could change the
relationship between compressibility and informativeness. Some work has shown
that the ``dimensionality'' of neural representations can change systematically
with task complexity, even in response to an identical
stimulus~\citep{MackEtal20}. Taken together, we expect that the way we have
defined informativeness in this paper, and the specific dataset we examined,
are likely to have influenced our findings. While we see our approach as a
reasonable first step, we also suggest that future work should explore
alternative measures of informativeness and compressibility, and should examine
how these measures vary across different tasks and datasets.
